---
output: github_document
---

```{r opts, include = FALSE}
knitr::opts_chunk$set(message = FALSE, 
                      warning = FALSE,
                      eval = FALSE, 
                      echo = TRUE)
```


[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)
[![CRAN](https://www.r-pkg.org/badges/version/spatsoc)](https://cran.r-project.org/package=spatsoc)
[![codecov](https://codecov.io/gl/robit.a/spatsoc/branch/master/graph/badge.svg)](https://codecov.io/gl/robit.a/spatsoc)
[![pipeline status](https://gitlab.com/robit.a/spatsoc/badges/master/pipeline.svg)](https://gitlab.com/robit.a/spatsoc/commits/master)



# spatsoc 
spatsoc is an R package for detecting spatial and temporal groups in GPS relocations. It can be used to convert GPS relocations to gambit-of-the-group format to build proximity-based social networks. In addition, the randomization function provides data-stream randomization methods suitable for GPS data. 

See below for installation and basic usage. 

For more detailed function descriptions, see the vignettes:

* [Introduction to spatsoc](https://spatsoc.gitlab.io/articles/intro-spatsoc.html)

## Installation
`spatsoc` depends on `rgeos` and requires [GEOS](https://trac.osgeo.org/geos/) installed on the system. 

* Debian/Ubuntu: `apt-get install libgeos-dev`
* Arch: `pacman -S geos`
* Fedora: `dnf install geos geos-devel`
* Mac: `brew install geos`
* Windows: see [here](https://trac.osgeo.org/osgeo4w/)


```{r install}
# Development version
devtools::install_git('https://gitlab.com/robit.a/spatsoc.git')

# CRAN
install.packages('spatsoc')
```

## Usage
### Import package, read data
```{r library}
library(spatsoc)
library(data.table)
DT <- fread(system.file("extdata", "DT.csv", package = "spatsoc"))
DT[, datetime := as.POSIXct(datetime, tz = 'UTC')]
```
### Temporal grouping

#### group_times
```{r groupTimes}
group_times(
  DT, 
  datetime = 'datetime', 
  threshold = '5 minutes'
)
```

### Spatial grouping
#### group_pts

```{r groupPts}
group_pts(
  DT,
  threshold = 5,
  id = 'ID',
  coords = c('X', 'Y'),
  timegroup = 'timegroup'
)
```


#### group_lines

```{r groupLines}
utm <-
  '+proj=utm +zone=36 +south +ellps=WGS84 +datum=WGS84 +units=m +no_defs'
  
group_times(
  DT, 
  datetime = 'datetime', 
  threshold = '1 day'
)

group_lines(
  DT,
  threshold = 50,
  projection = utm,
  id = 'ID',
  coords = c('X', 'Y'),
  timegroup = 'timegroup'
)
```


#### group_polys

```{r groupPolys}
utm <- '+proj=utm +zone=36 +south +ellps=WGS84 +datum=WGS84 +units=m +no_defs'

group_polys(
  DT,
  area = FALSE,
  'mcp',
  list(percent = 95),
  projection = utm,
  id = 'ID',
  coords = c('X', 'Y')
)
  
areaDT <- group_polys(
  DT,
  area = TRUE,
  'mcp',
  list(percent = 95),
  projection = utm,
  id = 'ID',
  coords = c('X', 'Y')
)
```

