---
title: "FAQ about spatsoc"
author: "Alec Robitaille, Quinn Webber and Eric Vander Wal"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    number_sections: true
vignette: >
  %\VignetteIndexEntry{FAQ}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitropts, include = FALSE}
knitr::opts_chunk$set(message = FALSE, 
                      warning = FALSE,
                      eval = FALSE, 
                      echo = TRUE)
```

# 
## Don't I need to reassign to save the output?
(Almost) all functions in `spatsoc` use data.table's modify-by-reference to reduce recopying large datasets and improve performance. The expection is `group_polys(area = TRUE)`. 



## Why does a function print the result, but columns aren't added to my DT?

Check that your data.table has columns allocated (with `data.table::truelength`) and if not, use `data.table::setDT`. This can happen if you are reading your data from `RDS` or `RData` files [1](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-faq.html#reading-data.table-from-rds-or-rdata-file). 

```{r alloc}
if (truelength(DT) == 0) {
  setDT(DT)
}
```

