---
title: "Frequently asked questions about spatsoc"
author: "Alec Robitaille, Quinn Webber and Eric Vander Wal"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    number_sections: true
    toc: true
vignette: >
  %\VignetteIndexEntry{FAQ}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitropts, include = FALSE}
knitr::opts_chunk$set(message = FALSE, 
                      warning = FALSE,
                      eval = FALSE, 
                      echo = TRUE)
```



* how do i use randomizations? 
* how do i build a graph with the outputs?
* my timegroup changes when i reorder
* GEOS
* gbi
* passing to igraph/asnipe
* not a moving window
* direction of proportion


# Usage
`spatsoc` is designed to work in two steps: temporal followed by spatial grouping. Considering your specific study species and system, determine a relevant temporal and spatial grouping threshold. This may be 5 minutes and 50 meters or 2 days and 100 meters or any other thresholds - that decision is left to you.  

```{r}
# Load packages
library(spatsoc)
library(data.table)

# Read data as a data.table
DT <- fread(system.file("extdata", "DT.csv", package = "spatsoc"))

# Cast datetime column to POSIXct
DT[, datetime := as.POSIXct(datetime)]

# Temporal groups
group_times(DT, datetime = 'datetime', threshold = '5 minutes')

# Spatial groups
group_pts(
  DT,
  threshold = 50,
  id = 'ID',
  coords = c('X', 'Y'),
  timegroup = 'timegroup'
)
```



## `group_times`

### `datetime` format
The `group_times` function expects either one column: `POSIXct` or two columns: `IDate` and `ITime`. These are provided to the function using the names of the column in the input data. 

`group_times(DT, datetime = 'datetime', threshold = '5 minutes')`

`group_times(DT, datetime = c('idate', 'itime'), threshold = '5 minutes')`


### `threshold` recommendations
The `threshold` provided to `group_times` should be related to the fix rate of the input dataset or to the specific study system and species. If relocations are recorded every two hours, a `threshold = '2 hours'` will group all rows to the nearest two hour group (10am, 12pm, 2pm, 4pm, ...). This, however, means that the relocations can be up to one hour apart from each other. Picking a smaller threshold, e.g.: `threshold = '15 minutes'` may be more relevant. The flexibility of `spatsoc`'s threshold argument means the user must carefully consider what threshold is reasonable to their specific system. 

### Limitations of `threshold`
Because there is no 


# Package design
`spatsoc` leverages `data.table` to modify-by-reference and iteratively work on subsets of the input data. 

## Don't I need to reassign to save the output?

(Almost) all functions in `spatsoc` use data.table's modify-by-reference to reduce recopying large datasets and improve performance. The exception is `group_polys(area = TRUE)`. 


## Why does a function print the result, but columns aren't added to my DT?

Check that your data.table has columns allocated (with `data.table::truelength`) and if not, use `data.table::setDT`. This can happen if you are reading your data from `RDS` or `RData` files.  [See here.](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-faq.html#reading-data.table-from-rds-or-rdata-file)

```{r alloc}
if (truelength(DT) == 0) {
  setDT(DT)
}
# then go to spatsoc
group_times(DT, datetime = 'datetime', threshold = '5 minutes')
```

