---
title: "Introduction to spatsoc"
author: "Alec Robitaille"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
library(ggplot2); library(data.table); library(spatsoc)
data(locs)
utm <- '+proj=utm +zone=21 ellps=WGS84'

sublocs <- locs[EASTING < 640000 & (ID %in% c('ind23', 'ind11', 'ind15'))][order(idate, itime)]
sublocs[, datetime := as.POSIXct(paste(idate, itime))]

ggplot(sublocs) + 
  geom_point(aes(EASTING, NORTHING, color = ID, group = ID))# + guides(color = FALSE)


ggplot(fortify(rgeos::gBuffer(BuildPts(sublocs, utm), 50, byid = FALSE))) + 
  geom_polygon(aes(long, lat, group = group, color = id, fill = id), alpha = 0.2)# +


GroupPts

ggplot(sublocs) + 
  geom_path(aes(EASTING, NORTHING, color = ID, group = ID))# + guides(color = FALSE)

hrs <- BuildHRs(hrType = 'mcp', DT = sublocs, projection = utm)
ggplot(fortify(hrs)) + 
  geom_polygon(aes(long, lat, group = group, color = id, fill = id), alpha = 0.2)# +
  # geom_point(data = sublocs, aes(EASTING, NORTHING, color = ID, group = ID)) + # + guides(color = FALSE)
  # geom_path(data = sublocs, aes(EASTING, NORTHING, color = ID, group = ID))# + guides(color = FALSE)

```











The `spatsoc` package provides functionality for analysing animal location data in time and space to identify potential interactions among individuals or social groups. 

The package contains a set of `Group*` functions that are used for identifying groups from input data depending on spatial location and (optionally) time. In addition, the package contains functions for determining the nearest neighbor to each individual, as well as the mean pairwise distance between a set of individuals (both optionally across time). 

## Functions
* Grouping 
    * `GroupPts`  
    * `GroupLines`  
    * `GroupHRs` / `GroupPolys` 
    * `GroupClusters` 
    * `CalcCentroids`
    * `GroupTime`
   
* Mean pairwise distance
    * `PairwiseDist`

* Nearest neighbor
    * `Nearest`

* Randomization of IDs within groups
    * `Randomizations`
    
* Helper functions
    * `BuildPts`
    * `BuildLines`
    * `BuildHRs`
    * `BuildClusters`
    * `RoundTime`

## Grouping
Let's read in `spatsoc`'s example data.

```{r}
library(spatsoc); library(data.table)
data(locs)
locs[1:5]
```

<!-- note this input/example data is just for now.. -->
The data consist of relocations of `r locs[, uniqueN(ID)]` individuals over `r locs[, max(idate) - min(idate)]` days. Relocations were recorded using GPS collars with fixes obtained every two hours. Using these data, we can compare the various grouping methods available in `spatsoc`. 

### `GroupPts`
The `GroupPts` function compares the relocations of all individuals at each time step in the input data and groups those within a specified buffer distance. 
Before grouping, we will ensure that the time field we intend to group on is in the correct format. Because GPS location fixes across animals are not 
necessarily obtained at the exact same time, we provide the function `GroupTimes` to find temporal groups within a specified time threshold. 

<!-- all below needs updating since GroupTimes replaces RoundTime -->

```{r}
locs[minute(itime) > 5, .N]
locs[minute(itime) > 5]
```

If we want to group individuals in time and space, it is clear that these 20 relocations may not be relevant to those recorded ~40-50 minutes earlier. We will round the times to the nearest hour using `RoundTime` and returning them `as.POSIXct` datetimes if a date field is provided. Alternatively, if the temporal scale of your input data is different, time fields can be rounded using different round units such as `'15 minutes'` or `'4 hours'`. <!-- for example this time -->

```{r, eval = FALSE}
locs[, roundDateTime := RoundTime(.SD, 'itime', roundUnit = '1 hour', dateField = 'idate')]
```

<!-- Finally, the data preparation is complete by using the grouppts to call helper function
above roundtime to group times to on nearest hour-->

We can then use the `GroupPts` function to find overlapping spatial location of buffered points. Here, we buffer each point by 50m to match temporal and spatial overlap of potential interation or social group.  

```{r, eval = FALSE}
GroupPts(locs, 50, timeField = 'roundDateTime', 
         projection = '+proj=utm +zone=21 ellps=WGS84', 
         idField = 'ID', coordFields = c('EASTING', 'NORTHING'))

```


<!-- ?QW when would best period be for social interactions -->
<!---
The basic methodology is as follows:

1. Group locations by overlapping or matching time. 
1. Group locations 
1. *Optionally* round time fields.
-->
## Notes
Package dependencies for `spatsoc` are <!-- list all and note details --> `sp`, `rgeos`, and `data.table`. `data.table` provides efficient methods for manipulating large (or small) datasets. As a result, input `DT` for all `spatsoc` functions must be a `data.table` and if it isn't, you can simply use `setDT(df)` to convert it by reference. 

In addition, since the `rgeos` package is used in most functions (notably `GroupPts` and `GroupLines`) the input `DT`'s coordinate system is important. `rgeos` expects planar coordinates and this requirement is carried forward for `spatsoc`. 
