---
title: "Introduction to spatsoc"
author: "Alec Robitaille, Quinn Webber and Eric Vander Wal"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to spatsoc}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r knitropts, include = FALSE}
knitr::opts_chunk$set(message = TRUE, 
                      warning = FALSE,
                      eval = FALSE, 
                      echo = FALSE)
```

The `spatsoc` package provides functionality for analyzing animal relocation data in time and space to identify potential interactions among individuals and build gambit-of-the-group data for constructing social networks. 

The package contains grouping functions that are used for identifying spatially and temporally explicit groups from input data. In addition, we provide a function for randomizing individual identifiers within groups, designed to test whether social networks generated from animal relocation data were based on non-random social proximity among individuals.

The functions were developed for application across animal relocation data, for example, proximity based social network analyses and spatial and temporal clustering of points.

## Grouping

```{r libs, eval = TRUE, include = FALSE}
library(spatsoc); library(data.table)
DT <- fread(system.file("extdata", "DT.csv", package = "spatsoc"))[ID %in% c('H', 'I', 'J')]
DT[, datetime := as.POSIXct(datetime, tz = 'UTC')]
```

`spatsoc` expects a `data.table` for all of its functions. If you have a `data.frame`, you can use `data.table::setDT()` to convert it by reference. If your data is a CSV, you can use `data.table::fread()` to import it as a `data.table`. 

The data consist of relocations of `r DT[, uniqueN(ID)]` individuals over `r DT[, max(yday(datetime)) - min(yday(datetime))]` days. Using these data, we can compare the various grouping methods available in `spatsoc`.


```{r, echo = TRUE}
## Load packages
library(spatsoc)
library(data.table)

## Read in spatsoc's example data
DT <- fread(system.file("extdata", "DT.csv", package = "spatsoc"))

## Subset to only individuals H, I, and J
DT <- DT[ID %in% c('H', 'I', 'J')]

## Cast character column 'datetime' as POSIXct
DT[, datetime := as.POSIXct(datetime,
                            tz = 'UTC')]
```

```{r, eval = TRUE}
knitr::kable(DT[, .SD[1:2], ID][order(datetime)])
```


### `group_times`
The `group_times` function is used to group relocations temporally. It is flexible to a threshold provided in units of minutes, hours or days. Since GPS fixes taken at regular intervals have some level of variability, we will provide a time threshold (`threshold`), to consider all fixes within this threshold taken at the same time. Alternatively, we may want to understand different scales of grouping, perhaps daily movement trajectories or seasonal home range overlap. 

```{r groupmins, echo = TRUE}
group_times(DT, datetime = 'datetime', threshold = '5 minutes')
```

```{r tableSetUp, eval = TRUE}
nRows <- 9
```

```{r tabgroupmins, eval = TRUE}
knitr::kable(
  group_times(DT, threshold = '5 minutes', datetime = 'datetime')[, 
    .(ID, X, Y, datetime, minutes, timegroup)][
      order(datetime)][
        1:nRows])
```


A message is returned when `group_times` is run again on the same `DT`, as the columns already exist in the input `DT` and will be overwritten. 

```{r grouphours, echo = TRUE}
group_times(DT, datetime = 'datetime', threshold = '2 hours')
```

```{r tabgrouphours, eval = TRUE}
knitr::kable(
  group_times(DT, threshold = '2 hours', datetime = 'datetime')[, 
    .(ID, X, Y, datetime, hours, timegroup)][
      order(datetime)][
        1:nRows])
```

```{r groupdays, echo = TRUE}
group_times(DT, datetime = 'datetime', threshold = '5 days')
```

```{r tabgroupdays, eval = TRUE}
knitr::kable(
  group_times(DT, threshold = '5 days', datetime = 'datetime')[, .SD[sample(.N, 3)], by = .(timegroup, block)][order(datetime)][
        1:nRows, .(ID, X, Y, datetime, block, timegroup)])
```


### `group_pts`
The `group_pts` function compares the relocations of all individuals in each timegroup and groups individuals based on a distance threshold provided by the user.

```{r grouppts, echo = TRUE}
group_times(DT = DT, datetime = 'datetime', threshold = '15 minutes')
group_pts(DT, threshold = 50, id = 'ID', coords = c('X', 'Y'), timegroup = 'timegroup')
```

```{r fakegrouppts, eval = TRUE}
DT <- group_times(DT = DT, datetime = 'datetime', 
                     threshold = '15 minutes')
DT <- group_pts(
    DT = DT,
    threshold = 50, id = 'ID', coords = c('X', 'Y'),
    timegroup = 'timegroup')

knitr::kable(
  DT[
      between(group,  771, 774)][order(timegroup)][
        1:nRows, .(ID, X, Y, timegroup, group)]
)
```


### `group_lines`
The `group_lines` function groups individuals whose trajectories intersect in a specified time interval. This represents a coarser grouping method than `group_pts` which can help understand shared space at daily, weekly or other temporal resolutions.

```{r fakegrouplines, echo = TRUE}
utm <- '+proj=utm +zone=21 ellps=WGS84'
group_times(DT = DT, datetime = 'datetime', threshold = '1 day')
group_lines(DT, threshold = 50, projection = utm, 
            id = 'ID', coords = c('X', 'Y'),
            timegroup = 'timegroup', sortBy = 'datetime')
```

```{r grouplines, eval = TRUE}
utm <- '+proj=utm +zone=21 ellps=WGS84'

DT <- group_times(DT = DT, datetime = 'datetime', 
                threshold = '1 day')
DT <- group_lines(DT,
                  threshold = 50, projection = utm, 
                  id = 'ID', coords = c('X', 'Y'), 
                  sortBy = 'datetime', timegroup = 'timegroup')
knitr::kable(
  unique(DT[, .(ID, timegroup, group)])[, .SD[1:3], ID][order(timegroup)]
)
```


### `group_polys`
The `group_polys` function groups individuals whose home ranges intersect. This represents the coarsest grouping method, to provide a measure of overlap across seasons, years or all available relocations. It can either return the proportion of home range area overlapping between individuals or simple groups. Home ranges are calculated using `adehabitatHR::kernelUD` or `adehabitatHR::mcp`. Alternatively, a `SpatialPolygonsDataFrame` can be input to the `spPolys` argument.

```{r fakegrouppolys, echo = TRUE}
utm <- '+proj=utm +zone=21 ellps=WGS84'
group_times(DT = DT, datetime = 'datetime', threshold = '8 days')
group_polys(DT = DT, area = TRUE, hrType = 'mcp',
           hrParams = list('percent' = 95),
           projection = utm,
           coords = c('X', 'Y'), id = 'ID')
```

```{r grouppolys, eval = TRUE}
utm <- '+proj=utm +zone=21 ellps=WGS84'
DT <- group_times(DT = DT, datetime = 'datetime', 
                threshold = '8 days')
knitr::kable(
  data.frame(group_polys(
    DT, 
    area = TRUE, hrType = 'mcp',
           hrParams = list('percent' = 95),
           projection = utm,
           coords = c('X', 'Y'), id = 'ID')[
             , .(ID1, ID2, area, proportion)])
)
```


## Notes
Package dependencies for `spatsoc` are `sp`, `rgeos`, `igraph`, `adehabitatHR` and `data.table`. `data.table` provides efficient methods for manipulating large (or small) datasets. As a result, input `DT` for all `spatsoc` functions must be a `data.table` and if it isn't, you can simply use `data.table::setDT(df)` to convert it by reference. 

In addition, since the `rgeos` package is used in most functions (`group_lines` and `group_polys`) the input `DT`'s coordinate system is important. `rgeos` expects planar coordinates and this requirement is carried forward for `spatsoc`. Since `rgeos` is used, system dependencies include `GEOS`. 
