---
title: "Introduction to spatsoc"
author: "Alec Robitaille"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The `spatsoc` package aims to provide functionality for analysing animal location data in time and space to find potential interactions or social groups. 

The package contains a set of `Group*` functions that are used for finding groups in input data depending on spatial location and (optionally) time. In addition, the package contains functions for determining the nearest neighbor to each individual, as well as the mean pairwise distance between a set of individuals (both optionally across time). 

## Functions
* Grouping 
    * `GroupPts`  
    * `GroupLines`  
    * `GroupHRs` / `GroupPolys` 
    * `GroupClusters` 
    * `CalcCentroids`
    * `GroupTime`
   
* Mean pairwise distance
    * `PairwiseDist`

* Nearest neighbor
    * `Nearest`

* Randomization of IDs within groups
    * `Randomizations`
    
* Helper functions
    * `BuildPts`
    * `BuildLines`
    * `BuildHRs`
    * `BuildClusters`
    * `RoundTime`

## Grouping
Let's read in `spatsoc`'s example data.

```{r}
library(spatsoc); library(data.table)
data(locs)
locs[1:5]
```

The data consists of the location of `r locs[, uniqueN(ID)]` individuals over `r locs[, max(idate) - min(idate)]` days. The location was recorded using GPS collars with two-hour fixes. Using this data, we can compare the various grouping methods available in `spatsoc`. 

### `GroupPts`
The `GroupPts` function compares the location of all individuals at each time step in the input data and groups those within a provided distance. 

Before grouping, we will ensure that the time field we intend to group on is suitable. First, we will check how many locations are taken more than 5 minutes off of the nearest hour. 

```{r}
locs[minute(itime) > 5, .N]
locs[minute(itime) > 5]
```

If we want to group individuals in time and space, it is clear that these 20 locations may not be relevant to those recorded ~40-50 minutes earlier. We will round the times to the nearest hour using `RoundTime` and returning them `as.POSIXct` datetimes if a date field is provided. Alternatively, if the temporal scale of your input data is different, time fields can be rounded using different round units such as `'15 minutes'` or `'4 hours'`. 

```{r, eval = FALSE}
locs[, roundDateTime := RoundTime(.SD, 'itime', roundUnit = '1 hour', dateField = 'idate')]
```

<!-- Finally, the data preparation is complete by using the grouppts to call helper function
above roundtime to group times to on nearest hour-->

We can then use the `GroupPts` function to find overlapping spatial location of buffered points. Here, we buffer each point by 50m to match temporal and spatial overlap of potential interation or social group.  

```{r, eval = FALSE}
GroupPts(locs, 50, timeField = 'roundDateTime', 
         projection = '+proj=utm +zone=21 ellps=WGS84', 
         idField = 'ID', coordFields = c('EASTING', 'NORTHING'))

```


<!-- ?QW when would best period be for social interactions -->
<!---
The basic methodology is as follows:

1. Group locations by overlapping or matching time. 
1. Group locations 
1. *Optionally* round time fields.
-->
## Notes
Package dependencies for `spatsoc` are primarily `sp`, `rgeos`, and `data.table`. `data.table` provides efficient methods for manipulating large (or small) datasets. As a result, input `DT` for all `spatsoc` functions must be a `data.table` and if it isn't, you can simply use `setDT(df)` to convert it by reference. 

In addition, since the `rgeos` package is used in most functions (notably `GroupPts` and `GroupLines`) the input `DT`'s coordinate system is important. `rgeos` expects planar coordinates and this requirement is carried forward for `spatsoc`. 
