---
title: "Using spatsoc in social network analysis"
author: "Alec Robitaille, Quinn Webber and Eric Vander Wal"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    number_sections: false
    toc: false
vignette: >
  %\VignetteIndexEntry{Using spatsoc for social network analysis}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r knitropts, include = FALSE}
knitr::opts_chunk$set(message = TRUE, 
                      warning = FALSE,
                      eval = FALSE, 
                      echo = TRUE)
```


`spatsoc` can be used in social network analysis to generate gambit of the group format data from GPS relocation data and perform data stream randomization. Gambit of the group format data can be used to generate social networks. 

Gambit of the group format data is generated using the grouping functions:

* `group_times`
* `group_pts`
* `group_lines`
* `group_polys`

Data stream randomization is performed using the `randomization` function. 

# Generate network
spatsoc provides users with one temporal (`group_times`) and three spatial (`group_pts`, `group_lines`, `group_polys`) functions to generate gambit of the group data from GPS relocations. Users can consider spatial grouping at 3 different scales combined with an appropriate temporal grouping threshold. The gambit of the group data is then used to generate a group by individual matrix and build the network. 


## 1. Import packages and prepare data

<!-- data.table, posixct -->



```{r}
# Load packages
library(spatsoc)
library(data.table)

# Read data as a data.table
DT <- fread(system.file("extdata", "DT.csv", package = "spatsoc"))

# Cast datetime column to POSIXct
DT[, datetime := as.POSIXct(datetime)]
```


Next, we will group relocations temporally with `group_times` and spatially with one of `group_pts`, `group_lines`, `group_polys`. Note: these are mutually exclusive, only select one spatial grouping function at a time. 

<!-- based... maybe scope? scale? or just group_pts-->
## 2. a) Point based grouping 

Depending on species and study system, relevant temporal and spatial grouping thresholds are used. In this case, relocations within 5 minutes and 50 meters are grouped together. 

```{r}
# Temporal groups
group_times(DT, datetime = 'datetime', threshold = '5 minutes')

# Spatial groups
group_pts(
  DT,
  threshold = 50,
  id = 'ID',
  coords = c('X', 'Y'),
  timegroup = 'timegroup'
)

```

## 2. b) Line based grouping 

Longer temporal threshold to create, for example, daily trajectories that are optionally buffered. 

```{r}
# UTM zone for relocation
utm <- '+proj=utm +zone=36 +south +ellps=WGS84 +datum=WGS84 +units=m +no_defs'

# Group relocations by julian day
group_times(DT, datetime = 'datetime', threshold = '1 day')

# Group lines for each individual and julian day
group_lines(
  DT,
  threshold = 50,
  projection = utm,
  id = 'ID',
  coords = c('X', 'Y'),
  timegroup = 'timegroup',
  sortBy = 'datetime'
)
```


## 2. c) Polygon based grouping 

Longer temporal threshold to create seasonal, monthly, yearly home ranges and compare intersection or proportional overlap.

```{r}
# Proj4 string for example data
utm <- '+proj=utm +zone=36 +south +ellps=WGS84 +datum=WGS84 +units=m +no_defs'

# Option 1: area = FALSE and home range intersection 'group' column added to DT 
group_polys(
  DT,
  area = FALSE,
  hrType = 'mcp',
  hrParams = list(percent = 95),
  projection = utm,
  id = 'ID',
  coords = c('X', 'Y')
)

# Option 2: area = TRUE and results must be assigned to a new variable
#      data.table returned has ID 1, ID 2 and proportion and area overlap
areaDT <- group_polys(
  DT,
  area = TRUE,
  hrType = 'mcp',
  hrParams = list(percent = 95),
  projection = utm,
  id = 'ID',
  coords = c('X', 'Y')
)

```

# Data stream randomization
Three types of data stream randomization are provided by `spatsoc`'s `randomization` function:


* step: randomizes identities of relocations between individuals within each time step.
* daily: randomizes identities of relocations between individuals within each day.
* trajectory: randomizes daily trajectories within individuals ([Spiegel et al. 2016](https://besjournals.onlinelibrary.wiley.com/doi/abs/10.1111/2041-210X.12553)).

Note: if `iterations = 1`, columns are added to the `DT`. If `iterations` is greater than 1, the results must be assigned.  
As with spatial grouping functions, these methods are mutually exclusive. Pick one `type` and rebuild the network after randomization. 

<!-- multi year note?  -->


<!-- or.. step randomization -->
## 3. a) `type = 'step'`
`'step'` randomizes identities of relocations between individuals within each time step. The `datetime` argument expects an integer group created by `group_times`. 

The first example shows where `iterations = 1`. A column named 'randomID' is appended corresponding to randomly selected ID from IDs within each time step. 

If `iterations` is greater than 1, columns 'observed' and 'iteration' are included indicating observed rows and which iteration rows correspond to (where 0 is the observed).


```{r}
# Iterations = 1
randomizations(
   DT,
   type = 'step',
   id = 'ID',
   datetime = 'timegroup',
   iterations = 1,
   splitBy = 'yr'
)

```


## 3. b) `type = 'daily'`
`'daily'` randomizes identities of relocations between individuals within each day. The `datetime` argument expects a datetime `POSIXct` format column. 

* daily - `randomID` for each day and `jul` indicating julian day

```{r}
# Calculate year column to ensure randomization only occurs within years since data spans multiple years
DT[, yr := year(datetime)]

# splitBy = 'yr' to force randomization only within year
randDaily <- randomizations(
   DT,
   type = 'daily',
   id = 'ID',
   datetime = 'datetime',
   splitBy = 'yr',
   iterations = 20
)

```

## 3. c) `type = 'trajectory'`
`'trajectory'` randomizes daily trajectories within individuals ([Spiegel et al. 2016](https://besjournals.onlinelibrary.wiley.com/doi/abs/10.1111/2041-210X.12553)). The `datetime` argument expects a datetime `POSIXct` format column. 

* trajectory - a random date time ("random" prefixed to *datetime* argument), observed *jul* and *randomJul* indicating the random day relocations are swapped to.
* observed - if the rows represent the observed (TRUE/FALSE)
* iteration - iteration of rows (where 0 is the observed)

```{r}
# Calculate year column to ensure randomization only occurs within years since data spans multiple years
DT[, yr := year(datetime)]

randTraj <- randomizations(
   DT,
   type = 'trajectory',
   id = 'ID',
   datetime = 'datetime',
   splitBy = 'yr',
   iterations = 20
)
```


# Example
Here is an example of point based spatial grouping with `spatsoc`'s "Newfoundland Bog Cow" example data. 

The data has a relocation rate of 2 hours but we will use a temporal threshold of 5 minutes. <!-- It is important to note that the temporal window is fixed in terms of 24 hours AND EXPLAIN LIMITATIONS?--> In addition, we will use a spatial distance threshold of 50 meters to accommodate the size and behaviour of the study species. The combination of spatial and temporal thresholds means that any individuals within 50 meters of each other within 5 minutes will be assigned to the same group.

## Generate network

```{r}
# Load packages
library(spatsoc)
library(data.table)

# Read data as a data.table
DT <- fread(system.file("extdata", "DT.csv", package = "spatsoc"))

# Cast datetime column to POSIXct
DT[, datetime := as.POSIXct(datetime)]

# Year of row 
DT[, yr := year(datetime)]

# Temporal groups
group_times(DT, datetime = 'datetime', threshold = '5 minutes')

# Spatial groups
group_pts(
  DT,
  threshold = 50,
  id = 'ID',
  coords = c('X', 'Y'),
  timegroup = 'timegroup'
)
d <- dcast(DT, formula = group ~ ID, 
           fun.aggregate = length, value.var = 'group')

gbi <- as.matrix(d[, .SD, .SDcols = colnames(d)[!grepl('group', colnames(d))]])
rownames(gbi) <- d$group

library(asnipe)
net <- get_network(gbi, data_format = "GBI", association_index = "SRI")

randDT <-
  randomizations(
    DT,
    type = 'step',
    id = 'ID',
    datetime = 'timegroup',
    iterations = 500
  )

group_pts(
  randDT,
  threshold = 50,
  id = 'randomID',
  coords = c('X', 'Y'),
  timegroup = 'timegroup',
  splitBy = 'iteration'
)

```
