---
title: "Using spatsoc in social network analysis - edge list generating functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using spatsoc in social network analysis - edge list generating functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(spatsoc)
```


`spatsoc` can be used in social network analysis to generate edge lists and spatiotemporal groups from GPS relocation data, perform data stream randomization and generate group by individual matrices.


Edge lists are generated using either the `edge_dist` or the `edge_nn` function. 

Data stream randomization is performed using the `randomizations` function. 

Group by individual matrices are generated using the `get_gbi` function.

**Note**: The grouping functions and their application in social network analysis are further described in the vignette [Using spatsoc in social network analysis - grouping functions](http://spatsoc.robitalec.ca/articles/using-in-sna.html). 