---
title: "Using spatsoc in social network analysis"
author: "Alec Robitaille, Quinn Webber and Eric Vander Wal"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    number_sections: false
    toc: false
vignette: >
  %\VignetteIndexEntry{Using spatsoc for social network analysis}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r knitropts, include = FALSE}
knitr::opts_chunk$set(message = TRUE, 
                      warning = FALSE,
                      eval = FALSE, 
                      echo = TRUE)
```


`spatsoc` can be used in social network analysis to generate gambit of the group format data from GPS relocation data and perform data stream randomization. Gambit of the group format data can be used to generate social networks. 

Gambit of the group format data is generated using the grouping functions:

* `group_times`
* `group_pts`
* `group_lines`
* `group_polys`

Data stream randomization is performed using the `randomization` function. 

# Generate network
spatsoc provides users with one temporal (`group_times`) and three spatial (`group_pts`, `group_lines`, `group_polys`) functions to generate gambit of the group data from GPS relocations. Users can consider spatial grouping at 3 different scales combined with an appropriate temporal grouping threshold. The gambit of the group data is then used to generate a group by individual matrix and build the network. 


## 1. Import packages and prepare data
`spatsoc` expects a `data.table` for all `DT` arguments and date time columns to be formatted `POSIXct`. 

```{r}
# Load packages
library(spatsoc)
library(data.table)

# Read data as a data.table
DT <- fread(system.file("extdata", "DT.csv", package = "spatsoc"))

# Cast datetime column to POSIXct
DT[, datetime := as.POSIXct(datetime)]
```


Next, we will group relocations temporally with `group_times` and spatially with one of `group_pts`, `group_lines`, `group_polys`. Note: these are mutually exclusive, only select one spatial grouping function at a time. 

## 2. a) `group_pts` 

Point based grouping by calculating distance between relocations in each timegroup. Depending on species and study system, relevant temporal and spatial grouping thresholds are used. In this case, relocations within 5 minutes and 50 meters are grouped together. 

```{r}
# Temporal groups
group_times(DT, datetime = 'datetime', threshold = '5 minutes')

# Spatial groups
group_pts(
  DT,
  threshold = 50,
  id = 'ID',
  coords = c('X', 'Y'),
  timegroup = 'timegroup'
)

```

## 2. b) `group_lines`

Line based grouping by measuring intersection of, optionally buffered, trajectories for each individual in each timegroup. Longer temporal thresholds are used to measure, for example, intersecting daily trajectories. 

```{r}
# UTM zone for relocation
utm <- '+proj=utm +zone=36 +south +ellps=WGS84 +datum=WGS84 +units=m +no_defs'

# Group relocations by julian day
group_times(DT, datetime = 'datetime', threshold = '1 day')

# Group lines for each individual and julian day
group_lines(
  DT,
  threshold = 50,
  projection = utm,
  id = 'ID',
  coords = c('X', 'Y'),
  timegroup = 'timegroup',
  sortBy = 'datetime'
)
```


## 2. c) `group_polys`

Polygon based grouping by generating home ranges using `adehabitatHR` and measuring intersection or proportional overlap. Longer temporal thresholds are used to create seasonal, monthly, yearly home ranges.

```{r}
# Proj4 string for example data
utm <- '+proj=utm +zone=36 +south +ellps=WGS84 +datum=WGS84 +units=m +no_defs'

# Option 1: area = FALSE and home range intersection 'group' column added to DT 
group_polys(
  DT,
  area = FALSE,
  hrType = 'mcp',
  hrParams = list(percent = 95),
  projection = utm,
  id = 'ID',
  coords = c('X', 'Y')
)

# Option 2: area = TRUE and results must be assigned to a new variable
#      data.table returned has ID 1, ID 2 and proportion and area overlap
areaDT <- group_polys(
  DT,
  area = TRUE,
  hrType = 'mcp',
  hrParams = list(percent = 95),
  projection = utm,
  id = 'ID',
  coords = c('X', 'Y')
)

```

# Data stream randomization
Three types of data stream randomization are provided by `spatsoc`'s `randomization` function:


* step: randomizes identities of relocations between individuals within each time step.
* daily: randomizes identities of relocations between individuals within each day.
* trajectory: randomizes daily trajectories within individuals ([Spiegel et al. 2016](https://besjournals.onlinelibrary.wiley.com/doi/abs/10.1111/2041-210X.12553)).

Note: if `iterations = 1`, columns are added to the `DT`. If `iterations` is greater than 1, the results must be assigned. In addition, if `iterations` is greater than 1, columns 'observed' and 'iteration' are returned indicating observed rows and which iteration rows correspond to (where 0 is the observed). <!-- rowID -->

As with spatial grouping functions, these methods are mutually exclusive. Pick one `type` and rebuild the network after randomization. 

## 3. a) `type = 'step'`
`'step'` randomizes identities of relocations between individuals within each time step. The `datetime` argument expects an integer group created by `group_times`. 

This example shows where `iterations = 1`. One column is returned when `type = 'step'`:

* 'randomID' - randomly selected ID from IDs within each time step

```{r}
# Iterations = 1 therefore the results are appended to DT
randomizations(
   DT,
   type = 'step',
   id = 'ID',
   datetime = 'timegroup',
   iterations = 1,
   splitBy = 'yr'
)
```


## 3. b) `type = 'daily'`
`'daily'` randomizes identities of relocations between individuals within each day. The `datetime` argument expects a datetime `POSIXct` format column. 

Two column is returned when `type = 'daily'`:

* 'randomID' - randomly selected ID for each day
* 'jul' - julian day

Since `iterations` is greater than 1, columns 'observed' and 'iteration' are returned indicating observed rows and which iteration rows correspond to (where 0 is the observed).

```{r}
# Calculate year column to ensure randomization only occurs within years since data spans multiple years
DT[, yr := year(datetime)]

# splitBy = 'yr' to force randomization only within year
# Iterations > 1, results must be reassigned
randDaily <- randomizations(
   DT,
   type = 'daily',
   id = 'ID',
   datetime = 'datetime',
   splitBy = 'yr',
   iterations = 20
)
```

## 3. c) `type = 'trajectory'`
`'trajectory'` randomizes daily trajectories within individuals ([Spiegel et al. 2016](https://besjournals.onlinelibrary.wiley.com/doi/abs/10.1111/2041-210X.12553)). The `datetime` argument expects a datetime `POSIXct` format column. 

Three column is returned when `type = 'trajectory'`:

* random date time ("random" prefixed to *datetime* argument)
* 'jul' - observed julian day
* 'randomJul' - random julian day relocations are swapped to

Since `iterations` is greater than 1, columns 'observed' and 'iteration' are returned indicating observed rows and which iteration rows correspond to (where 0 is the observed).

```{r}
# Calculate year column to ensure randomization only occurs within years since data spans multiple years
DT[, yr := year(datetime)]

randTraj <- randomizations(
   DT,
   type = 'trajectory',
   id = 'ID',
   datetime = 'datetime',
   splitBy = 'yr',
   iterations = 20
)
```


<!-- example with NL bog cow -->
